---
title: "HW1 Peer Assessment"
output:
  word_document: default
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Part A. ANOVA

Additional Material: ANOVA tutorial

https://datascienceplus.com/one-way-anova-in-r/

Jet lag is a common problem for people traveling across multiple time zones, but people can gradually adjust to the new time zone since the exposure of the shifted light schedule to their eyes can resets the internal circadian rhythm in a process called “phase shift”. Campbell and Murphy (1998) in a highly controversial study reported that the human circadian clock can also be reset by only exposing the back of the knee to light, with some hailing this as a major discovery and others challenging aspects of the experimental design. The table below is taken from a later experiment by Wright and Czeisler (2002) that re-examined the phenomenon. The new experiment measured circadian rhythm through the daily cycle of melatonin production in 22 subjects randomly assigned to one of three light treatments. Subjects were woken from sleep and for three hours were exposed to bright lights applied to the eyes only, to the knees only or to neither (control group). The effects of treatment to the circadian rhythm were measured two days later by the magnitude of phase shift (measured in hours) in each subject’s daily cycle of melatonin production. A negative measurement indicates a delay in melatonin production, a predicted effect of light treatment, while a positive number indicates an advance.

Raw data of phase shift, in hours, for the circadian rhythm experiment

|Treatment|Phase Shift (hr)                            |
|:--------|:-------------------------------------------|
|Control  |0.53, 0.36, 0.20, -0.37, -0.60, -0.64, -0.68, -1.27|
|Knees    |0.73, 0.31, 0.03, -0.29, -0.56, -0.96, -1.61       |
|Eyes     |-0.78, -0.86, -1.35, -1.48, -1.52, -2.04, -2.83    |

## Question A1 - 3 pts

Consider the following incomplete R output:

|Source|Df |Sum of Squares|Mean Squares|F-statistics|p-value|
|:----:|:-:|:------------:|:----------:|:----------:|:-----:|
|Treatments|?|?|3.6122|?|0.004|
|Error|?|9.415|?| | |
|TOTAL|?|?| | | |

Fill in the missing values in the analysis of the variance table.

```{r data setup}
##Read data
sleep <- read.csv("sleep.csv",sep = ",",header=TRUE,fileEncoding="UTF-8-BOM")
##Fit a linear regression model
sleep_aov <- aov(sleep$Phase~sleep$Treatment)
summary(sleep_aov)
##Totals
Total_SSq=7.224+9.415
Total_Df=2+19
cat("TOTAL:","Df:",Total_Df,"SSq:",Total_SSq)
```

## Question A2 - 3 pts

Use $\mu_1$, $\mu_2$, and $\mu_3$  as notation for the three mean parameters and define these parameters clearly based on the context of the topic above. Find the estimates of these parameters.
```{r}
mu_names<- list("mu_1","mu_2","mu_3")
cat("\tmu_1","mu_2","mu_3\n",sep="\t")
#print(\n)
#tapply(sleep$Phase, sleep$Treatment, mean)
model.tables(sleep_aov,type="means")
cat("mu_1 (Control)","mu_2 (Eyes)","mu_3 (Knees)",sep=" ")
cat("\n\nmu_1: Average magnitude of phase shift (in hours) in daily cycle of melatonin production for control group.\n","\nmu_2: Average magnitude of phase shift (in hours) in daily cycle of melatonin production for eye group.\n","\nmu_3: Average magnitude of phase shift (in hours) in daily cycle of melatonin production for knee group.\n",sep=" ")
```
## Question A3 - 5 pts

Use the ANOVA table in Question A1 to answer the following questions:

a. **1 pts** Write the null hypothesis of the ANOVA $F$-test, $H_0$
```{r}
cat("H_0:","mu_1 =","mu_2 =","mu_3",sep=" ")
```
b. **1 pts** Write the alternative hypothesis of the ANOVA $F$-test, $H_A$
```{r}
cat("H_A:","At least two means from the groups are not equal")
```

c. **1 pts** Fill in the blanks for the degrees of freedom of the ANOVA $F$-test statistic:   $F$(____, _____)

```{r}
k <-3
N <- 22

f1<-k-1
f2<-N-k
sprintf("ANOVA F-test statistic:F(%s, %s)",f1,f2)
```

d. **1 pts** What is the p-value of the ANOVA $F$-test?
```{r}
print("P-value: 0.00447")
```

e. **1 pts** According the the results of the ANOVA $F$-test, does light treatment affect phase shift?  Use an $\alpha$-level of 0.05.

```{r}
print("Since the p-value of 0.0045 is less than the alpha level of 0.05, we can reject the null hypothesis. This indicates that the differences in the means for the groups are statistically significant and that the light treat does affect phase shift.")
```

# Part B. Simple Linear Regression

We are going to use regression analysis to estimate the performance of CPUs based on the maximum number of channels in the CPU.  This data set comes from the UCI Machine Learning Repository.

The data file includes the following columns:

* *vendor*: vendor of the CPU
* *chmax*: maximum channels in the CPU
* *performance*: published relative performance of the CPU

The data is in the file "machine.csv". To read the data in `R`, save the file in your working directory (make sure you have changed the directory if different from the R working directory) and read the data using the `R` function `read.csv()`.

```{r}
# Read in the data
data = read.csv("machine.csv", head = TRUE, sep = ",")
# Show the first few rows of data
head(data, 3)
```

## Question B1: Exploratory Data Analysis - 9 pts

a. **3 pts** Use a scatter plot to describe the relationship between CPU performance and the maximum number of channels. Describe the general trend (direction and form). Include plots and R-code used.

```{r}
perf<-data$performance
chan<- data$chmax
plot(perf~chan)
```

From this scatterplot, we can see that the form of the scatterplot is non linear in a positive direction, but seems to have  moderately high strength with a few outliers.



b. **3 pts** What is the value of the correlation coefficient between _performance_ and _chmax_? Please interpret the strength of the correlation based on the correlation coefficient.

```{r}
cat("The correlation coefficient is:",cor(perf,chan))
print("This correlation coefficient indicates the correlation to be of high strength.")
```

c. **2 pts** Based on this exploratory analysis, would you recommend a simple linear regression model for the relationship?

No since the scatterplot is non linear, may be more useful on data with transformations with a stronger correlation.

d. **1 pts** Based on the analysis above, would you pursue a transformation of the data? *Do not transform the data.*

Yes, I would pursue a transformation of the data


## Question B2: Fitting the Simple Linear Regression Model - 11 pts

Fit a linear regression model, named *model1*, to evaluate the relationship between performance and the maximum number of channels. *Do not transform the data.* The function you should use in R is:

```{r}
# Your code here...
model1 = lm(performance ~ chmax, data)
summary(model1)
```

a. **3 pts** What are the model parameters and what are their estimates?  

The model parameters are the:

intercept Parameter
beta_0: 37.2252

slope parameters
beta_1: 3.7441

Variance of error terms:
128.3^2 = 16,460.9


b. **2 pts** Write down the estimated simple linear regression equation.

Performance = 37.2252 + 3.7441*(Max Channels)

c. **2 pts** Interpret the estimated value of the $\beta_1$ parameter in the context of the problem.

With a positive beta_1 of 3.7441, the predictive variable (Max number of Channels) and response variable (performance) have a direct relationship.This indicates for each increase in incremental channels of 3.7441, performance will increase by one unit.

d. **2 pts** Find a 95% confidence interval for the $\beta_1$ parameter. Is $\beta_1$ statistically significant at this level?

```{r}
confint(model1,level=0.95)
```
Yes, beta_1 is statistically significant at this level.

e. **2 pts** Is $\beta_1$ statistically significantly positive at an $\alpha$-level of 0.01?  What is the approximate p-value of this test?
```{r}
tval <- 10.938
1-pt(tval,207)

```
Yes because the p value is 0.

## Question B3: Checking the Assumptions of the Model - 8 pts

Create and interpret the following graphs with respect to the assumptions of the linear regression model. In other words, comment on whether there are any apparent departures from the assumptions of the linear regression model. Make sure that you state the model assumptions and assess each one.  Each graph may be used to assess one or more model assumptions.

a. **2 pts** Scatterplot of the data with *chmax* on the x-axis and *performance* on the y-axis

```{r}
plot(perf~chan)
```

**Model Assumption(s) it checks: Linearity**
**Interpretation: shape indicates Non-linear, linearity does not hold**

b. **3 pts** Residual plot - a plot of the residuals, $\hat\epsilon_i$, versus the fitted values, $\hat{y}_i$

```{r}
plot(model1$fitted,model1$residuals)

```

**Model Assumption(s) it checks: constant variance and assumption of uncorrelated errors**

**Interpretation: Since scatter plot is not random around 0 line, constant variance does not hold**

c. **3 pts** Histogram and q-q plot of the residuals

```{r}
#library("car")
res<-model1$residuals
histogram(model1$residuals)
qqPlot(model1$residuals)
```

**Model Assumption(s) it checks: normality**

**Interpretation: Skew and gap in histogram as well as departures in tail of q-q plot indicates Normality does not hold**


## Question B4: Improving the Fit - 10 pts

a. **2 pts** Use a Box-Cox transformation (`boxCox()`) to find the optimal $\lambda$ value rounded to the nearest half integer.  What transformation of the response, if any, does it suggest to perform?

```{r}
boxCox(model1)

```
Optimal lambda is around 0, indicating we use a normal logarithmic transformation.

b. **2 pts** Create a linear regression model, named *model2*, that uses the log transformed *performance* as the response, and the log transformed *chmax* as the predictor. Note: The variable *chmax* has a couple of zero values which will cause problems when taking the natural log. Please add one to the predictor before taking the natural log of it

```{r}
model2 = lm(log(performance) ~ log(chmax+1), data)
summary(model2)
```

e. **2 pts** Compare the R-squared values of *model1* and *model2*.  Did the transformation improve the explanatory power of the model?

Yes, the R-squared increased from.36 to 0.407 and the transformation improved the explanatory power.

c. **4 pts** Similar to Question B3, assess and interpret all model assumptions of *model2*.  A model is considered a good fit if all assumptions hold. Based on your interpretation of the model assumptions, is *model2* a good fit?

```{r}
plot(log(perf) ~ log(chan+1))
plot(model2)

```
Looking at at the log of performance on one axis and the log of max channel on another, the scatter plot holds a linear shape, upholding the linear assumption. 

The scatter plot with the fitted and residuals also look randomly positioned around the zero line indicating constant variance holds.

The QQ plot with the fitted and residuals follow a linear pattern with minimal departures in tail, indicating Normality holds.
## Question B5: Prediction - 3 pts

Suppose we are interested in predicting CPU performance when `chmax = 128`.  Please make a prediction using both *model1* and *model2* and provide the 95% prediction interval of each prediction on the original scale of the response, *performance*. What observations can you make about the result in the context of the problem?

```{r}
#Model1
#Performance1 = 37.2252 + 3.7441*(128)
#Model2
#Performance2 = 2.47655 + 0.64819*(128)
#Performance1
#exp(Performance2-1)
newppp = data.frame(chmax = c(128))
print("Prediction intervals")
predict(model1,newppp,interval=c("prediction"))
predict(model2,exp(newppp-1),interval=c("prediction"))
```

# Part C. ANOVA - 8 pts

We are going to continue using the CPU data set to analyse various vendors in the data set.  There are over 20 vendors in the data set.  To simplify the task, we are going to limit our analysis to three vendors, specifically, honeywell, hp, and nas.  The code to filter for those vendors is provided below.

```{r}
# Filter for honeywell, hp, and nas
data2 = data[data$vendor %in% c("honeywell", "hp", "nas"), ]
data2$vendor = factor(data2$vendor)
```

1. **2 pts** Using `data2`, create a boxplot of *performance* and *vendor*, with *performance* on the vertical axis.  Interpret the plots.  

```{r}
boxplot(data2$performance,data2$vendor)

```

2. **3 pts** Perform an ANOVA F-test on the means of the three vendors.  Using an $\alpha$-level of 0.05, can we reject the null hypothesis that the means of the three vendors are equal?  Please interpret.

```{r}
a <-aov(data2$performance~data2$vendor)
summary(a)
```

3. **3 pts** Perform a Tukey pairwise comparison between the three vendors. Using an $\alpha$-level of 0.05, which means are statistically significantly different from each other?

```{r}
# Your code here...

```
